PAMELA: an annotation-based Java Modeling Framework

### 2021-02-10 ResponseLetter.tex

% Title:    A LaTeX Template For Responses To a Referees' Reports
% Author:   Petr Zemek <s3rvac@gmail.com>
% Homepage: https://blog.petrzemek.net/2016/07/17/latex-template-for-responses-to-referees-reports/
% License:  CC BY 4.0 (https://creativecommons.org/licenses/by/4.0/)
\documentclass[10pt]{article}

% Allow Unicode input (alternatively, you can use XeLaTeX or LuaLaTeX)
\usepackage[utf8]{inputenc}
\usepackage{hyperref}
\usepackage{xcolor}

\usepackage{microtype,xparse,tcolorbox}
\newenvironment{reviewer-comment}{}{}
\tcbuselibrary{skins}
\tcolorboxenvironment{reviewer-comment }{empty,
  left = 1em, top = 1ex, bottom = 1ex,
  borderline west = {2pt} {0pt} {black!20},
}
\ExplSyntaxOn
\NewDocumentEnvironment {response} { +m O{black!20} } {
  \IfValueT {#1} {
    \begin{reviewer-comment~}
      \setlength\parindent{2em}
      \noindent
      \ttfamily #1
    \end{reviewer-comment~}
  }
  \par\noindent\ignorespaces\color{blue}
} { \bigskip\par }

\NewDocumentCommand \Reviewer { m } {
  \section*{Comments~by~Reviewer~#1}
}

\NewDocumentCommand \ToolComments { m } {
  \section*{Tool~Comments~by~Reviewer~#1}
}

\ExplSyntaxOff
\AtBeginDocument{\maketitle\thispagestyle{empty}\noindent}

% You can get probably get rid of these definitions:
\newcommand\meta[1]{$\langle\hbox{#1}\rangle$}
\newcommand\PaperTitle[1]{``\textit{#1}''}

\title{Response letter to Manuscript SCICO-D-20-00176 entitled "PAMELA: an Annotation-Based Java Modeling Framework"}
%\author{Author1 \and Author2 \and Author3}
\date{\today}

\begin{document}

Dear Editor,

\bigskip
Thanks for your message of January 9th, 2021, in which you suggested preparing a major revision of our manuscript  \textbf{PAMELA: an Annotation-Based Java Modeling Framework}.

\bigskip
We are grateful to the reviewers for their interest, effort, and suggestions. We detail in the following pages how we have addressed the reviewers' concerns; we are confident that the paper is now stronger, and hope that you will agree. We have organized the responses to the comments on a per-reviewer basis, focusing, as suggested, on:

\begin{itemize}
\item Improving the readability of the manuscript. 
\item Improving tool documentation and examples.
\item Provide additional experiments \& evaluation.
\end{itemize}

Moreover, the paper has been thoroughly revised to fix minor issues and to give the additional clarifications required by the reviewers. Please let us know if you need any further information or clarification.

\bigskip
Sincerely yours,

\bigskip
Sylvain, Guillaume, Caine, Joel, Jean-Christophe, Salvador and Antoine

\pagebreak

\Reviewer{\#1}
\begin{response}{They then briefly discuss implementation details, in a single page; and then discuss 'industrial use cases and experiments'.

Unfortunately, the latter leaves much to be desired, as much of the text is spent on the context/tools in which the tool was used, but not on how it was used, nor with any quantification on how extensive its use was, or what the performance of the resulting constellation is.
}
  Response to the first comment.
\end{response}
\begin{response}{Also lacking is a more thorough comparison to existing work; your related work section gives a very shallow comparison to other approaches..}
  Response to the second comment.
\end{response}

\begin{response}{What would also be interesting is a general discussion on performance of your tool, gains by using your tool, ...
}
  Response to n-ieme comment.
\end{response}

\begin{response}{Your reference list is incomplete and inconsistent. Formod or Formosa? When were URLs accessed (add date). DOIs are missing. Ref. 5 should have the ISoLA track name and editors added. Ref. 7 is missing publisher, journal, conference. Inconsistent use of first names, first names + initials, first initial.}
  Response to n-ieme comment.
\end{response}

\begin{response}{The paper highlights include "Contract programming with assertions [remove the plural] checking at runtime", yet that is not mentioned in the paper at all.
}
  Response to n-ieme comment.
\end{response}

\begin{response}{That also brings up the question: what about correctness of your code? Any evidence for that?}
  Response to n-ieme comment.
\end{response}

\begin{response}{p10 l32+: very little detail or quantification on the tool's code base. Surely you can say more about its structure, size of components, ... other interesting and relevant observations on the codebase?}
  Response to n-ieme comment.
\end{response}

\begin{response}{p10 l32+: very little detail or quantification on the tool's code base. Surely you can say more about its structure, size of components, ... other interesting and relevant observations on the codebase?}
  Response to n-ieme comment.
\end{response}

\ToolComments{\#1}

\begin{response}{gives "git clone git@github.com:openflexo-team/pamela.git", which (even when using my own username) does not work. However, "git clone https://github.com/openflexo-team/pamela.git" immediately does. Why not state the latter?
}
  Response to n-ieme comment.
\end{response}

\begin{response}{how to actually build the environment is not stated for this option.}
  Response to n-ieme comment.
\end{response}

\begin{response}{Run tests Unit tests located in src/test/java in pamela-core are a good starting point to discover the framework." No further information is given. Apparently, the user has to figure out him/herself how to run the tests. Documentation is lacking here.
}
  Response to n-ieme comment.
\end{response}

\begin{response}{Make your first model." And then what?
}
  Response to n-ieme comment.
\end{response}

\begin{response}{Summarizing: the *getting started* documentation is incomplete and certainly not suitable for getting started. The page "https://pamela.openflexo.org/SNAPSHOT/example.html" gives a basic example, but the relation between that and the getting started page is not made explicit, nor do the two in their current shape provide a user with a proper getting started, allowing him/her to at least (1) install the framework, (2) create a basic example, (3) use the framework with that basic example. This needs to be improved before an evaluation of the software can be completed.
}
  Response to n-ieme comment.
\end{response}

\pagebreak

\Reviewer{\#2}
\begin{response}{scenarios are missing. As a developer I don't know in which cases it is relevant to use this tool.}
  Response to the first comment.
\end{response}

\begin{response}{how to use the tool is not describe properly. It is not clear if the developers need to write all the annotations. If an application is built with PAMELA, what is the link between the interfaces and the concrete classes. Indeed, at one point, code corresponding to the behaviour will be needed; and not only the accessors to navigate the structure of the model.}
  Response to n-ieme comment.
\end{response}

\begin{response}{how does such "code" evolve with time? Authors have used PAMELA for years, how such applications resist with time?}
  Response to n-ieme comment.
\end{response}

\begin{response}{section 2 is too abstract to be very useful}
  Response to n-ieme comment.
\end{response}

\begin{response}{p6 "according to PAMELA semantics" but it is never explained.}
  Response to n-ieme comment.
\end{response}

\begin{response}{the partially implemented class concept is not clear.}
  Response to n-ieme comment.
\end{response}

\begin{response}{the additional features seem relevant, but they are not described. Either the authors describe them a bit more or they remove the section.}
  Response to n-ieme comment.
\end{response}

\begin{response}{in section 3.4 it is not clear who writes what and at which step of the development process.}
  Response to n-ieme comment.
\end{response}

\begin{response}{in section 5, the use of PAMELA in the two projects is not clear.}
  Response to n-ieme comment.
\end{response}

\begin{response}{references are missing. For example Kermeta done, Platypus (from Alain Plantec) done, work of Glenn Cavarle and others from the model @runtime community.}
  Response to n-ieme comment.
\end{response}

\begin{response}{in section 6, the authors write: "PAMELA does not focus in reflecting and manipulating running systems but in: 1) extending their behaviours; and 2) maintaining a continuous cycle of metamodeling... These two points are not clear previously in the paper.
}
  Response to n-ieme comment.
\end{response}

\ToolComments{\#2}

\begin{response}{The link in the paper is not clickable and copy-and-paste does not work (because of latex?) Link found: https://github.com/openflexo-team/pamela}
  Response to n-ieme comment.
\end{response}

\begin{response}{Installation
The readme of the GitHub page should link to the installation page found in https://pamela.openflexo.org. It is unclear we must visit this web site to install the tool.
}
  Response to n-ieme comment.
\end{response}

\begin{response}{
What I have done

Create a new project in Intellij Idea with gradle
Add the line found in \url{https://pamela.openflexo.org/SNAPSHOT/getting_started.html}

- Test 1: Create a model with only one entity without error

- Test 2: Create the example of the paper

--->Bug: in Listing2, the authors execute Line9 ".setName" but the method does not exists
Else, the example works.

}
  Response to n-ieme comment.
\end{response}

\end{document}

### 2021-02-10 implementation.tex
Adding response letter template

\subsection{PAMELA code base figures}

\todo{Surely you can say more about its structure, size of components, ... other interesting and relevant observations on the codebase?}

>>>>

\subsection{PAMELA code base metrics}

\sg{

PAMELA implementation is modularized.

\begin{itemize}
    \vspace{-0.2cm}\item Core implementation (\texttt{pamela-core}) provides all base features for PAMELA. It contains 20k lines of code involving 184 classes. This code is covered with unit test (6k lines of code, 112 classes). Reached code coverage is about 66\%.
    \vspace{-0.2cm}\item \texttt{pamela-security-patterns} is an add-on library containing some security-pattern implementations.
    \vspace{-0.2cm}\item \texttt{pamela-perf-tests} contains benchmarking tools whose purpose is to quantifiate PAMELA performances. 
\end{itemize}

}
%\todo{Surely you can say more about its structure, size of components, ... other interesting and relevant observations on the codebase?}

\subsection{On performance issues}

\sg{
Compared to a basic POJO (Plain Object Java Object) implementation, use of PAMELA involves a significant CPU and memory-footprint overhead, because of the partially-interpretated nature for some model features.

A workbench\footnote {\url{https://github.com/openflexo-team/pamela/tree/1.6.1/pamela-perf-tests}} has been developped to help measure that overhead. We use a base model composed of four entities defining 4 to 5 properties (single and multiple properties), and we instantiate that tree model (1010101 objects created and initialized to default value). From this simple model, two java implementations have been derived using code generation. The first implementation was made using fully implemented java classes (POJO), while the second one is fully interpretated (it uses PAMELA framework and defines only interfaces and API methods).

Unsurprisingly, we quantifiate a significant CPU overhead (from x20). This is due to the fact that \texttt{MethodHandler} mechanism is fully interpretated and contains many hooks, and cannot be compared to a fully-compiled and optimized Java dynamic binding. As well, memory footprint is increased from a x5 factor.

These conclusions must be moderated by the fact that this overhead only applies to a very small portion of code (mostly property accessors methods). A "real-world" model implementation generally involves a bigger "business code" part, which is generally the first CPU-time consumer. Using PAMELA won't increase CPU use on "business code" if this code is implemented with plain Java code (which is generally the case). As we will see in next section, PAMELA has been successfully applied to many industrial projects, where no major performance issues were reported.

Moreover, comparison between the two implementations should be tempered by the fact that PAMELA implementation offers many functional features which are not present in the base implementation (such as undo/redo, clipboard management, runtime monitoring and weaving, etc...).

}

### 2021-02-10 pamelaProcess.tex
Adding response letter template

We advocate for a strong coupling between model and source-code, to give architects and developers a way to both interact during the whole development cycle. PAMELA is an annotation-based Java modeling framework providing a smooth integration between model and code, without code generation nor externalized model serialization. The idea is to avoid separation between modeling and code to facilitate consistency management and to /avoid round-trip issues. Figure \ref{fig:PamelaVision} summarizes the PAMELA architecture.

@@

Coupling model and code into the same artifact opens new ways of programming. The classical (metadata enabled) programming process relies on \emph{programmers} that produce code reusing pre-existing modeling concepts. These concepts are implemented by \emph{modelers} that provide the right annotations the programmers use. This is, for instance, the process followed by Jakarta EE (JEE) developers reusing JEE specific annotations. The evolution rhythm between models and code is low. This programming way is still possible with PAMELA, but we allow the ability to reach a high evolution rhythm when the programmer also becomes the modeler. In fact, when a pattern, an abstraction, a generalization is identified by the programmer, s/he can use PAMELA to develop and capitalize on this abstraction by increasing PAMELA metamodel.

@@

>>>>

We advocate for a strong coupling between model and source-code, to give architects and developers a way to both interact during the whole development cycle. PAMELA is an annotation-based Java modeling framework providing a smooth integration between model and code, without code generation nor externalized model serialization. \sg{Instead of generating the code, the API (mostly Java interfaces with abstract method declarations) is locally executed.}. The idea is to avoid separation between modeling and code to facilitate consistency management and to /avoid round-trip issues. Figure \ref{fig:PamelaVision} summarizes the PAMELA architecture.

@@

Coupling model and code into the same artifact opens new ways of programming. The classical (metadata enabled) programming process relies on \emph{programmers} that produce code reusing pre-existing modeling concepts. These concepts are implemented by \emph{modelers} that provide the right annotations the programmers use. This is, for instance, the process followed by Jakarta EE (JEE) developers reusing JEE specific annotations. The evolution rythm between models and code is low. This programming way is still possible with PAMELA, but we allow the ability to reach a high evolution rythm when the programmer also becomes the modeler. In fact, when a pattern, an abstraction, a generalization is identified by the programmer, s/he can use PAMELA to develop and capitalize on this abstraction by increasing PAMELA metamodel.

@@

\sg{
Here are various conceivable scenarios for PAMELA use:
\begin{itemize}
    \vspace{-0.2cm}\item \textbf{Programming use} : programmers reuse existing annotations and write model and code at the same time (the modelers and programmers share the same artifacts : the Java code). This scenario includes the case where the model (made by the modelers) is pre-existing.
    \vspace{-0.2cm}\item \textbf{Reengineering} : programmers start from an existing base of code (legacy) and refactor it while replacing existing code by abstract method declarations in Java interface, reducing risks of errors.
    \vspace{-0.2cm}\item \textbf{Aspect-oriented programming} : programmers may use or redefine "patterns" (eg Security Patterns) which offers code weaving at runtime, as well as runtime monitoring.
    \vspace{-0.2cm}\item \textbf{Advanced programmers use} : programmers may extends PAMELA with their own annotations, their own implementations, their own patterns, etc...
\end{itemize}
}
